Ed. Psych. 964: HLM Assignment 3
===============================================================================

_Tristan Mahr, April 2014_


Codebook
-------------------------------------------------------------------------------

* `id`: subject ID
* `attit`: attitudes toward deviance (composite score). Larger values imply more tolerance toward deviant behaviors by peers (such as cheating, stealing five dollars, etc.).
* `expo`: exposure to delinquent peers (composite score). Larger values imply more frequent contacts with delinquent peers.
* `age`: age of subject
* `age11`: `age` minus 11
* `age13`: `age` minus 13
* `age11s`: `age11` squared
* `age13s`: `age13` squared
* `ind1`--`ind5`: measurement indicator matrix for HLM7.  Same for `age`, `age11`, `age13`.
* `female`: 1 if female, otherwise 0
* `minority`: 1 if minority, otherwise 0
* `income`: family income categories. Larger values imply higher family income.


Exercises
-------------------------------------------------------------------------------

```{r}
library(dplyr)
library(ggplot2)
library(lme4)
library(nlme)
d <- read.csv("nys.csv")
d <- select(d, -expo, -ind1, -ind2, -ind3, -ind4, -ind5)
head(d)

alpha_blue <- adjustcolor("steelblue", alpha.f = .25)
ggplot(d, aes(x = age, y = attit, group = id)) + 
  stat_smooth(method = "lm", se = FALSE, color = alpha_blue) + 
  stat_smooth(mapping = aes(group = NULL), method = "lm", size = 1.5, se = FALSE, color = "black")

ggplot(d, aes(x = age, y = attit, group = age)) + geom_boxplot()

```





**1\. Fit a linear growth model to these data allowing for random intercepts and slopes using `age11` as the level-one predictor, and `attit` as the outcome. Of the various types of level-1 covariance structures that can be applied using the HLM software, which appears to be the best for this model? Justify your answer.**


```{r}
# Decide on the random effects structure
m_rim <- lme(attit ~ age11, random = ~ 1 | id, data = d)
summary(m_rim)
m_rsm <-update(m_rim, random = ~ age11 | id, data = d)
summary(m_rsm)
anova(m_rim, m_rsm)
```


```{r}
# Look at the covariance structure
VarCorr(m_rsm)




# example code to follow...

tau00 <- as.numeric(VarCorr(RSM)[1, 1])
tau11 <- as.numeric(VarCorr(RSM)[2, 1])
tau10 <- as.numeric(VarCorr(RSM)[2, 3]) * sqrt(tau00 * tau11)
(TAU <- matrix(c(tau00, tau10, tau10, tau11), ncol = 2))

#         [,1]      [,2]
#[1,] 1236.425 -178.0690
#[2,] -178.069  107.2485


(sigma <- summary(RSM)$sigma ^ 2)
(design <- cbind(rep(1, nT), seq(0, nT - 1)))

# design matrix for "random" effects: note that the design matrix is a vector of ones in RIM

#       [,1] [,2]
# [1,]    1    0
# [2,]    1    1
# [3,]    1    2
# [4,]    1    3
 
(TAU.nT <- design %*% TAU %*% t(design))
#          [,1]      [,2]     [,3]      [,4]
#[1,] 1236.4253 1058.3563 880.2872  702.2182
#[2,] 1058.3563  987.5357 916.7152  845.8946
#[3,]  880.2872  916.7152 953.1431  989.5711
#[4,]  702.2182  845.8946 989.5711 1133.2476

(SIGMA.nT <- summary(RSM)$sigma ^ 2 * diag(nT))

#          [,1]     [,2]     [,3]     [,4]
# [1,] 159.4775   0.0000   0.0000   0.0000
# [2,]   0.0000 159.4775   0.0000   0.0000
# [3,]   0.0000   0.0000 159.4775   0.0000
# [4,]   0.0000   0.0000   0.0000 159.4775


(RSM.cov <- TAU.nT + SIGMA.nT)
#          [,1]      [,2]      [,3]      [,4]
#[1,] 1395.9028 1058.3563  880.2872  702.2182
#[2,] 1058.3563 1147.0132  916.7152  845.8946
#[3,]  880.2872  916.7152 1112.6206  989.5711
#[4,]  702.2182  845.8946  989.5711 1292.7250


```







**2\. Fit a quadratic growth model using `age11` as a predictor under the homogeneous covariance structure. Allow all of the growth parameters to be random. Report and interpret the parameter estimates. Does it appear necessary to add a quadratic term to accurately model change in attitudes toward deviant behavior?**



**3\. Are `gender`, `minority`, or `income` significant predictors of the random intercepts, slopes, or quadratic terms? Interpret any significant findings.**