
Preliminaries
-------------------------------------------------------------------------------

### Codebook

* `id`: subject ID
* `attit`: attitudes toward deviance (composite score). Larger values imply more tolerance toward deviant behaviors by peers (such as cheating, stealing five dollars, etc.).
* `expo`: exposure to delinquent peers (composite score). Larger values imply more frequent contacts with delinquent peers.
* `age`: age of subject
* `age11`: `age` minus 11
* `age13`: `age` minus 13
* `age11s`: `age11` squared
* `age13s`: `age13` squared
* `ind1`--`ind5`: measurement indicator matrix for HLM7.  Same for `age`, `age11`, `age13`.
* `female`: 1 if female, otherwise 0
* `minority`: 1 if minority, otherwise 0
* `income`: family income categories. Larger values imply higher family income.




### First look at the data

#### Descriptives

```{r descriptives}
head(d)
as.data.frame(table(table(d$id), dnn = c("ObservationsPerSubject")))
d$age_group <- paste0("Age_", d$age)
d$gender <- ifelse(d$female, "female", "male")
d$minority_status <- ifelse(d$minority, "minority", "non-minority")
describe_briefly <- function(...) {
  summary <- describeBy(..., digits = 3, mat = TRUE)
  select(summary, -vars, -trimmed, -mad, -min, -max, -item)
}

# Attitude by age, gender, minority, income
describe_briefly(d$attit, group = list(Age = d$age_group))
describe_briefly(d$attit, group = list(d$age_group, d$gender))
describe_briefly(d$attit, group = list(d$age_group, d$minority_status))
describe_briefly(d$attit, group = list(d$income))
```

#### Exploratory plots

```{r exploratory plots, warning = FALSE, message = FALSE}
ggplot(d, aes(x = age, y = attit, group = age)) + geom_boxplot() + 
  facet_grid(gender ~ minority_status)
# Within-subject change
alpha_blue <- "#4682B440"
ggplot(d, aes(x = age, y = attit, group = id)) + 
  stat_smooth(method = "lm", se = FALSE, color = alpha_blue) + 
  stat_smooth(aes(group = NULL), method = "lm", size = 1, color = "black")
# Within-subject change by group
ggplot(d, aes(x = age, y = attit, color = gender, linetype = minority_status)) + 
  facet_grid( ~ gender) + 
  stat_smooth(aes(group = id), method = "lm", se = FALSE, size = .1) + 
  stat_smooth(method = "lm", size = 1, color = "blue")
```


#### Intra-class correlation

```{r ICC}
m_empty <- lme(attit ~ 1, random = ~ 1 | id, data = d, method = "ML")
summary(m_empty)
cov_matrix2(m_empty)
#  Tau / (Tau + Sigma_Squared)
GmeanRel(m_empty)$ICC
```
