### 1\. Fit a linear growth model to these data allowing for random intercepts and slopes using `age11` as the level-one predictor, and `attit` as the outcome. Of the various types of level-1 covariance structures, which appears to be the best for this model? Justify your answer.

First we look at the least and most parameterized models.

```{r}
# Independence model
m_indep <- gls(attit ~ age11, data = d, method = "ML")  
invisible(summary(m_indep))
occasions <- max(table(d$id))
sigma_squared(m_indep) * diag(occasions)
# Autocorrelation of the residuals
ACF(m_indep, form = ~ 1 | id, resType = "normalized")
# Unstructured model
m_unstruct <- gls(attit ~ age11, data = d, 
                  correlation = corSymm(form = ~ 1 | id), 
                  weights = varIdent(form = ~ 1 | age11), method = "ML")
# Much less autocorrelation
ACF(m_unstruct, resType = "normalized")
cov_matrix(m_unstruct)
```

Then we fit a random intercept (compound symmetry) and random-slope model, then try different variance structures on those models.

```{r random-effect models}
# Models with random effects
m_rim <- lme(attit ~ age11, random = ~ 1 | id, data = d, method = "ML")
m_rsm <- lme(attit ~ age11, random = ~ age11 | id, data = d, method = "ML")

# RIM + autocorrelated covariances
m_rim_ar1 <- update(m_rim, correlation = corAR1(form = ~ 1 | id))
invisible(summary(m_rim_ar1))
cov_matrix2(m_rim)
cov_matrix2(m_rim_ar1)
ACF(m_rim_ar1, resType = "normalized")
# RSM + heterogeneous variances
m_rsm_hetero <- update(m_rsm, weights = varIdent(form = ~ 1 | age11))
summary(m_rsm_hetero)
ACF(m_rsm_hetero, resType = "normalized")
anova(m_indep, m_rim, m_rim_ar1, m_rsm, m_rsm_hetero, m_unstruct)
```

The unstructured model provides the best fit; it certainly has the smallest AIC value. The model's fit comes at the cost of increased parameters and the added work of interpreting those parameters. The random-slope model with heterogeneous variances provides the next best AIC value, and it is also much more comprehensible. This model claims:

* Attitudes of 11 to 15 year olds change over time.
* Individuals vary randomly in their baseline attitude and in the rate at which attitude changes over time. These two types of random variation are related.
* Even when we take into account the above effects, the amount of random variation in attitude is different at each age level. 

Moreover, we have to consider the goal of the modeling. For this dataset, we want to describe how attitude within-subjects changes over time, and the random effects models are geared towards this task. 

> The choice between an `lme` model and a `gls` model should take into account more than just information criteria and likelihood tests. A mixed-effects model has a hierarchical structure which, in many applications, provides a more intuitive way of accounting for within-group dependency than the direct modeling of the marginal variance-covariance structure of the response in the `gls` approach. Furthermore, the mixed-effects estimation gives, as a byproduct, estimates for the random effects, which may be of interest in themselves. The `gls` model focuses on marginal inference and is more appealing when a hierarchical structure for the data is not believed to be present, or is not relevant in the analysis, and one is more interested in parameters associated with the error variance-covariance structure, as in time-series analysis and spatial statistics. (Pinheiro & Bates, 2000, pp 254--255, as quoted in slides by James H. Steiger, `http://www.statpower.net/Content/GCM/Lectures/SW07.pdf`)

On these grounds---i.e., the purpose and performance of the model---the random slope model with heterogeneous variances is preferable.
